Finite Property Convex Elicitation Notes

### 2019-01-25 raf-notes.tex

@@

\raf{NOTES from call w/ Bo on 8/13/2018:}
\begin{itemize}
\item We would like a very strong link function, which takes any report in the convex domain, and returns the \emph{full set} of optimal reports, of the original finite property, but this is too strong -- consider abstain and the line segment connecting corner and center vs corner, center, and opposite corner.
So we should at least show the weaker link function that takes any report in the convex domain and returns \emph{some} optimal report of the original property.
\item We can probably prove that if $L$ essentially elicits a finite property, then $L$ is FPLC on the convex hull of the embedded reports.
\end{itemize}

@@

\begin{definition}
  We say a loss $L:\reals^d \to \reals^\Y$ is \emph{finite-piecewise-linear and convex (FPLC)} if for all $y\in\Y$, the function $L(\cdot)_y$ is a piecewise-linear function with finitely many pieces, convex, and bounded from below.
\end{definition}

@@

\begin{lemma}\label{lem:fplc-directional-deriv}
  Let $L:\reals^d \to \reals^\Y$ be FPLC, and let $r\in\reals$ be a point of differentiability of $L$.
  Let $d^-(r)_y$ and $d^+(r)_y$ denote the left and right derivative, respectively, of $L(\cdot)_y$ at $r$, so that $d^-(r),d^+(r)\in\reals^\Y$.
  Then for any open interval $(a,b)$ containing $r$ where $L$ is differentiable, we have
$d^+(a) = d^-(r) = d^+(r) = d^-(b)$.
\end{lemma}

@@

\begin{proposition}\label{prop:fplc-trim-char}
  Let $L : \reals \to \reals^\Y$ be FPLC and let $\R$ be the finite set of its nondifferentiable points.
  Then $\trim(\Gamma_L) \equiv \Gamma_L\cap\R'$ where $\underline r$ is the smallest $r\in\R$ such that $d^+(r)_y > 0$ for some $y\in\Y$, $\overline r$ is the largest such that $d^-(r)_y < 0$ for some $y\in\Y$, and $\R' = \R \cap [\underline r, \overline r]$.
\end{proposition}
\begin{proof}
  Since $L$ is FPLC, for each $y\in\Y$ we can take $\R_y$ to be the finite set of nondifferentiable points of $L(\cdot)_y$, and then $\R = \cup_{y\in\Y} \R_y$.
  As $L$ is bounded from below, we must have $d^-(\min \R)_y \leq 0$ and $d^+(\max \R)_y \geq 0$ for all $y\in\Y$.
  Thus, $\underline r$ and $\overline r$ are well-defined.

@@

  First, we argue that reports outside $[\underline r, \overline r]$ are weakly dominated.
  Let $r < \underline r$ and let $r'$ be the smallest value in $[r,\underline r]\cap \R$, i.e.,\ the next largest nondifferentiable point, which exists as $L$ is FPLC.
  If $r \in \Gamma_L(p)$, then we must have $d^+(r)\cdot p = 0$, as $d^+(r)\cdot p > 0$ would contradict the definition of $\underline r$.
  Then by Lemma~\ref{lem:fplc-directional-deriv}, as $L$ is differentiable on $(r,r')$, we have $0 = d^+(r)\cdot p = d^-(r')\cdot p$.
  Now, if $r' < \underline r$, we again have $d^+(r')\cdot p = 0$, and so on by induction (recall that $\R$ is finite) until we conclude $d^-(\underline r) = 0$, meaning $\underline r \in \Gamma_L(p)$ as well.
  Similarly, all reports $r > \overline r$ are weakly dominated by $\overline r$.

@@

\begin{corollary}\label{cor:fplc-orderable}
  Every FPLC $L : \reals \to \reals^\Y$ essentially elicits a finite orderable property.
\end{corollary}
\begin{proof}
  From Proposition~\ref{prop:fplc-trim-char}, it suffices to show that $\Gamma_L\cap\R'$ is orderable.
  \raf{I need to reference some other stuff above about $\trim(\Gamma_L)$.}
  Let $\R' = \{r_1,\ldots,r_m\}$ with $r_1 < \cdots < r_m$.
  Recalling the optimality condition for $r_i \in \Gamma_L(p)$ in eq.~\eqref{eq:1d-optimality}, we see that
  \begin{equation*}\label{eq:fplc-level-set}
    \Gamma_{r_i} = \{ p \in \Delta_\Y : p\cdot d^-(r_i) \leq 0 \leq p\cdot d^+(r_{i}) \}~.
  \end{equation*}
  Thus, observing that $d^+(r_i) = d^-(r_{i+1})$ from Lemma~\ref{lem:fplc-directional-deriv}, for all $i$ the intersection $\Gamma_{r_i} \cap \Gamma_{r_{i+1}} = \{ p \in \Delta_\Y : p\cdot d^+(r_i) = 0\}$ is a hyperplane.
\end{proof}

\begin{theorem}\label{thm:fplc-orderable}
  A finite property is essentially convex elicitable in $1$ dimension if and only if it is orderable.
  Moreover, this remains true when restricting to piecewise linear loss functions.
\end{theorem}
\begin{proof}
  From Corollary~\ref{cor:fplc-orderable}, it remains to show two things: (i) \raf{general convex losses}, and (ii) a finite ordeable property is elicitable by a FPLC loss function.  
\end{proof}

@@

\begin{lemma}\label{lem:sum-fplc}
  The sum of FPLC functions $f_1,\ldots,f_k$ is itself FPLC.
\end{lemma}
\begin{proof}
  Math.
\end{proof}

\begin{definition}
  Let $f:\reals^d\to\reals$ be FPLC.
  A \emph{vertex} of $f$ is a vertex of the corresponding convex polytope.
\end{definition}

\begin{lemma}\label{lem:vertiex-subgrad}
  Let $f:\reals^d\to\reals$ be FPLC.
  For every non-vertex $x$ of $f$, there is a vertex $v$ such that $\partial f(x) \subseteq \partial f(v)$.
\end{lemma}
\begin{proof}
  \raf{I'm thinking of a proof in two parts.  1. lemma from convex polytopes saying that you can always get to a vertex while staying within the current face; 2. all points in a face but no other face (so the relative interior) share the same subgradient set, and at the relative boundary, a superset of that set.}
\end{proof}

\begin{proposition}\label{cor:fplc-finite}
  Every FPLC $L : \reals^d \to \reals^\Y$ essentially elicits a finite property.
\end{proposition}
\begin{proof}
  Let $f(r) = \sum_{y\in\Y} L(r)_y$ which is FPLC by Lemma~\ref{lem:sum-fplc}.
  \raf{Perhaps its worth getting into power diagrams here...}
  Let $\R$ be the vertices of the cell complex given by $f$, that is, the points at which $\dim\partial f(r) = d$.
  As $f$ is FPLC, $\R$ is finite.
  We claim that $L$ essentially elicits $\Gamma_L\cap\R$.

@@

Note that the definition of $\R'$ is a generalization of its definition as $[\underline r, \overline r]$ in Proposition~\ref{prop:fplc-trim-char}.
\begin{conjecture}
  Let $L:\reals^d \to\reals^\Y$ be FPLC, and $\R$ the vertices of $\sum_{y\in\Y} L(\cdot)_y$.
  Then $\trim(\Gamma_L) \equiv \Gamma_L\cap \R'$ where $\R' = \conv \{ \argmin_{r\in\reals^d} L(r)_y : y\in\Y\} \cap \R$.
\end{conjecture}

\begin{conjecture}
  Let $L:\reals^d \to\reals^\Y$ be FPLC, and $\R$ the vertices of $f(\cdot) = \sum_{y\in\Y} L(\cdot)_y$.
  Then for all $p\in\inter\Delta(\Y)$, the convex hull of $\Gamma_L(p)\cap\R$ is a face of the polytope corresponding to $f$.
\end{conjecture}
\begin{proof}
  \raf{Sketch:}
  The polytope of $f$ is the same as that of $p\cdot L(\cdot)$, since $p$ has full support.  As $p\cdot L$ must be flat between all of $\Gamma_L(p)$, those vertices are on the same face.  Moreover, all vertices on the minimal face containing $\Gamma_L(p)$ are optimal.
\end{proof}

>>>>

\newcommand{\prop}[1]{\mathsf{prop}[#1]}
\newcommand{\eliccvx}{\mathrm{elic}_\mathrm{cvx}}
\newcommand{\elicpoly}{\mathrm{elic}_\mathrm{pcvx}}
\newcommand{\elicembed}{\mathrm{elic}_\mathrm{embed}}

@@

\raf{NOTES from call w/ Bo on 8/13/2018:}
\begin{itemize}
\item We would like a very strong link function, which takes any report in the convex domain, and returns the \emph{full set} of optimal reports, of the original finite property, but this is too strong -- consider abstain and the line segment connecting corner and center vs corner, center, and opposite corner.
So we should at least show the weaker link function that takes any report in the convex domain and returns \emph{some} optimal report of the original property.
\item We can probably prove that if $L$ essentially elicits a finite property, then $L$ is polyhedral on the convex hull of the embedded reports.
\end{itemize}

@@

\begin{definition}
  We say a loss $L:\reals^d \to \reals^\Y$ is \emph{finite-piecewise-linear and convex (polyhedral)} if for all $y\in\Y$, the function $L(\cdot)_y$ is a piecewise-linear function with finitely many pieces, convex, and bounded from below.
\end{definition}

@@

\begin{lemma}\label{lem:fplc-directional-deriv}
  Let $L:\reals^d \to \reals^\Y$ be polyhedral, and let $r\in\reals$ be a point of differentiability of $L$.
  Let $d^-(r)_y$ and $d^+(r)_y$ denote the left and right derivative, respectively, of $L(\cdot)_y$ at $r$, so that $d^-(r),d^+(r)\in\reals^\Y$.
  Then for any open interval $(a,b)$ containing $r$ where $L$ is differentiable, we have
$d^+(a) = d^-(r) = d^+(r) = d^-(b)$.
\end{lemma}

@@

\begin{proposition}\label{prop:fplc-trim-char}
  Let $L : \reals \to \reals^\Y$ be polyhedral and let $\R$ be the finite set of its nondifferentiable points.
  Then $\trim(\Gamma_L) \equiv \Gamma_L\cap\R'$ where $\underline r$ is the smallest $r\in\R$ such that $d^+(r)_y > 0$ for some $y\in\Y$, $\overline r$ is the largest such that $d^-(r)_y < 0$ for some $y\in\Y$, and $\R' = \R \cap [\underline r, \overline r]$.
\end{proposition}
\begin{proof}
  Since $L$ is polyhedral, for each $y\in\Y$ we can take $\R_y$ to be the finite set of nondifferentiable points of $L(\cdot)_y$, and then $\R = \cup_{y\in\Y} \R_y$.
  As $L$ is bounded from below, we must have $d^-(\min \R)_y \leq 0$ and $d^+(\max \R)_y \geq 0$ for all $y\in\Y$.
  Thus, $\underline r$ and $\overline r$ are well-defined.

@@

  First, we argue that reports outside $[\underline r, \overline r]$ are weakly dominated.
  Let $r < \underline r$ and let $r'$ be the smallest value in $[r,\underline r]\cap \R$, i.e.,\ the next largest nondifferentiable point, which exists as $L$ is polyhedral.
  If $r \in \Gamma_L(p)$, then we must have $d^+(r)\cdot p = 0$, as $d^+(r)\cdot p > 0$ would contradict the definition of $\underline r$.
  Then by Lemma~\ref{lem:fplc-directional-deriv}, as $L$ is differentiable on $(r,r')$, we have $0 = d^+(r)\cdot p = d^-(r')\cdot p$.
  Now, if $r' < \underline r$, we again have $d^+(r')\cdot p = 0$, and so on by induction (recall that $\R$ is finite) until we conclude $d^-(\underline r) = 0$, meaning $\underline r \in \Gamma_L(p)$ as well.
  Similarly, all reports $r > \overline r$ are weakly dominated by $\overline r$.

@@

\begin{corollary}\label{cor:fplc-orderable}
  Every polyhedral $L : \reals \to \reals^\Y$ essentially elicits a finite orderable property.
\end{corollary}
\begin{proof}
  From Proposition~\ref{prop:fplc-trim-char}, it suffices to show that $\Gamma_L\cap\R'$ is orderable.
  \raf{I need to reference some other stuff above about $\trim(\Gamma_L)$.}
  Let $\R' = \{r_1,\ldots,r_m\}$ with $r_1 < \cdots < r_m$.
  Recalling the optimality condition for $r_i \in \Gamma_L(p)$ in eq.~\eqref{eq:1d-optimality}, we see that
  \begin{equation*}\label{eq:fplc-level-set}
    \Gamma_{r_i} = \{ p \in \Delta_\Y : p\cdot d^-(r_i) \leq 0 \leq p\cdot d^+(r_{i}) \}~.
  \end{equation*}
  Thus, observing that $d^+(r_i) = d^-(r_{i+1})$ from Lemma~\ref{lem:fplc-directional-deriv}, for all $i$ the intersection $\Gamma_{r_i} \cap \Gamma_{r_{i+1}} = \{ p \in \Delta_\Y : p\cdot d^+(r_i) = 0\}$ is a hyperplane.
\end{proof}

\begin{theorem}\label{thm:fplc-orderable}
  A finite property is essentially convex elicitable in $1$ dimension if and only if it is orderable.
  Moreover, this remains true when restricting to piecewise linear loss functions.
\end{theorem}
\begin{proof}
  From Corollary~\ref{cor:fplc-orderable}, it remains to show two things: (i) \raf{general convex losses}, and (ii) a finite ordeable property is elicitable by a polyhedral loss function.  
\end{proof}

\subsection{New stuff (as of 1/23/2019)}

\begin{definition}
  A property $\Gamma:\Delta_\Y\to\R$ is \emph{monotone} if there are maps $a:\R\to\reals^\Y$, $b:\R\to\reals^\Y$ and a total ordering $<$ of $\R$ such that the following two conditions hold.
  \begin{enumerate}
  \item For all $r\in\R$, we have $\Gamma_r = \{p\in\Delta_\Y : a(r) \cdot p \leq 0 \leq b(r) \cdot p\}$.
  \item For all $r < r'$, we have $b(r) \leq a(r')$ (component-wise).
  % \item For all $r,r'\in\R$ and $p\in\Gamma_{r'}\setminus\Gamma_r$, we have $b(r) \cdot p < 0 \implies r' > r$ and $a(r) \cdot p > 0 \implies r' < r$.
  \end{enumerate}
\end{definition}

\begin{proposition}
  A property $\Gamma:\Delta_\Y\toto\reals$ is elicitable by a convex loss $L : \reals \to \reals^\Y$ if and only if it is monotone.
\end{proposition}
\begin{proof}
  Given such an $L$ eliciting $\Gamma$, we let $a,b$ be defined by $a(r)_y = \partial_- L(r)_y$ and $b(r) = \partial_+ L(r)_y$, that is, the left and right derivatives of $L(\cdot)_y$ at $r$.
  Then $\partial L(r)_y = [a(r)_y,b(r)_y]$.
  We now have $r \in \prop{L}(p) \iff 0 \in \partial p\cdot L(r) \iff a(r)\cdot p \leq 0 \leq b(r) \cdot p$, showing the first condition.
  The second condition follows as the subgradients of $L$ are monotone functions.

  Now given a monotone $\Gamma$, let $L(r)_y = \int_{-\infty}^r b(r')_y dr'$... \raf{FILL IN LATER if we care; only really need the above direction}  
\end{proof}

\begin{lemma}
  If $\gamma$ is monotone, finite, elicitable, and non-redundant, it is orderable.
\end{lemma}

\begin{theorem}
  Let $\gamma$ be a finite elicitable non-redundant property.
  If $\gamma$ is indirectly elicitable via a convex $L:\reals\to\reals^\Y$, then $\gamma$ is 1-embeddable.
  In other words, if $\eliccvx(\gamma)=1$, then $\elicpoly(\gamma) = \elicembed(\gamma)=1$.
\end{theorem}

@@

\begin{lemma}\label{lem:sum-fplc}
  The sum of polyhedral functions $f_1,\ldots,f_k$ is itself polyhedral.
\end{lemma}
\begin{proof}
  Math.
\end{proof}

\begin{definition}
  Let $f:\reals^d\to\reals$ be polyhedral.
  A \emph{vertex} of $f$ is a vertex of the corresponding convex polytope.
\end{definition}

\begin{lemma}\label{lem:vertiex-subgrad}
  Let $f:\reals^d\to\reals$ be polyhedral.
  For every non-vertex $x$ of $f$, there is a vertex $v$ such that $\partial f(x) \subseteq \partial f(v)$.
\end{lemma}
\begin{proof}
  \raf{I'm thinking of a proof in two parts.  1. lemma from convex polytopes saying that you can always get to a vertex while staying within the current face; 2. all points in a face but no other face (so the relative interior) share the same subgradient set, and at the relative boundary, a superset of that set.}
\end{proof}

\begin{proposition}\label{cor:fplc-finite}
  Every polyhedral $L : \reals^d \to \reals^\Y$ essentially elicits a finite property.
\end{proposition}
\begin{proof}
  Let $f(r) = \sum_{y\in\Y} L(r)_y$ which is polyhedral by Lemma~\ref{lem:sum-fplc}.
  \raf{Perhaps its worth getting into power diagrams here...}
  Let $\R$ be the vertices of the cell complex given by $f$, that is, the points at which $\dim\partial f(r) = d$.
  As $f$ is polyhedral, $\R$ is finite.
  We claim that $L$ essentially elicits $\Gamma_L\cap\R$.

@@

Note that the definition of $\R'$ is a generalization of its definition as $[\underline r, \overline r]$ in Proposition~\ref{prop:fplc-trim-char}.
\begin{conjecture}
  Let $L:\reals^d \to\reals^\Y$ be polyhedral, and $\R$ the vertices of $\sum_{y\in\Y} L(\cdot)_y$.
  Then $\trim(\Gamma_L) \equiv \Gamma_L\cap \R'$ where $\R' = \conv \{ \argmin_{r\in\reals^d} L(r)_y : y\in\Y\} \cap \R$.
\end{conjecture}

\begin{conjecture}
  Let $L:\reals^d \to\reals^\Y$ be polyhedral, and $\R$ the vertices of $f(\cdot) = \sum_{y\in\Y} L(\cdot)_y$.
  Then for all $p\in\inter\Delta(\Y)$, the convex hull of $\Gamma_L(p)\cap\R$ is a face of the polytope corresponding to $f$.
\end{conjecture}
\begin{proof}
  \raf{Sketch:}
  The polytope of $f$ is the same as that of $p\cdot L(\cdot)$, since $p$ has full support.  As $p\cdot L$ must be flat between all of $\Gamma_L(p)$, those vertices are on the same face.  Moreover, all vertices on the minimal face containing $\Gamma_L(p)$ are optimal.
\end{proof}
